\section{Related Work}
\label{sec:related}

In this paper we explored two techniques for capture avoiding substitution that avoids renaming, for the purpose of implementing static name binding in languages with $\lambda$s.
The topic of evaluating $\lambda$ expressions has a long and rich history.
Summarizing it all is beyond the scope of this paper; for overviews see, e.g., the works of Barendregt~\cite{DBLP:journals/bsl/Barendregt97} or Cardone~and~Hindley~\cite{hindley-lambda-history}.
We discuss a few of the papers that are most closely related to the techniques we have described.

In their formalization of $\lambda$ calculus and type theory, McKinna~and~Pollack~\cite{DBLP:journals/jar/McKinnaP99} consider a system that uses named substitution without renaming, for a particular notion of open terms.
They consider a syntax that distinguishes two classes of names: \emph{parameters} and \emph{variables}.
\emph{Variable substitution} does not affect parameters, and \emph{parameter substitution} does not affect variables.
Their notion of variable substitution is defined for terms that are \emph{variable-closed}, but which may be \emph{parameter-open}.
Thus, by encoding free variables as parameters, their system can be used to compute with open terms.
However, syntactically distinguishing free variables this way seems to presupposes a static binding analysis.
The approach we discussed in \cref{sec:interpreting-open} does not presuppose such static analysis.

Our paper considers how to interpret open terms.
There exist several calculi in the literature for evaluating open terms.
Accatolli~and~Guerrieri~\cite{DBLP:conf/aplas/AccattoliG16} gives an overview of several of these calculi for \emph{open call-by-value}, which is the class of languages that the interpreters in \cref{sec:interpretation}~and\cref{sec:shifting} interpret.
In their paper, Accatolli~and~Guerrieri focus on the meta-theory of these calculi.
For their meta-theoretical study they rely on an unspecified notion of capture-avoiding substitution.
In this paper, we explore how to define such capture-avoiding substitution in a way that does not perform renaming.
While the meta-theory of Berkling-Fehr substitution has been studied~\cite{berkling1982amodification}, the meta-theory of the substitution strategy in~\cref{sec:interpreting-open} remains an open question.

\section{Conclusion}
\label{sec:conclusion}

We have discussed two techniques for implementing capture avoiding substitution in definitional interpreters in a way that does not require renaming of bound variables.
One of the techniques relies on a coarse-grained but simple discipline for closing terms, is known to not support interpretation strategies that evaluate under binders, and has (to the best of our knowledge) not been studied meta-theoretically.
The other technique is an existing technique from the literature.
While this technique is less efficient, it is more fine-grained and so does is not subject to the same limitations as the first technique.
It also has a well-established meta-theory. 
Neither of the two techniques seem to be widely known or at least not widely applied.
With this work, we hope to increase awareness of these techniques.

%%% Local Variables:
%%% reftex-default-bibliography: ("../references.bib")
%%% End:
